---
title: "Health and economic impacts of extreme weather-related events"
subtitle: "A case study of the United States between 1996 and 2011"
author: "Victoria Mestre Runge"
date: "21 agost de 2020"
output: 
        github_document:
           html_preview: TRUE
           toc: TRUE
           toc_depth: 2 
           fig_width: 10
           fig_height: 10
           dev: png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the US National Oceanic and Atmospheric Administration's (NOAA) storm database which tracks characteristics of major storms and weather events in the US including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The analysis on the data revealed that tornadoes are the most harmful in terms of public health. The total injuries and fatalities produced by tornadoes far exceeded other weather/storm events and had the greatest overall health impact. In terms of monetary damage and loss, analysis on the data showed that floods are by far the most costly event, affecting both property and crops and resulting in the largest economic impact.

## Data processing

#### Data and documentation

Data come in a file compressed via the bzip2 which can be found in the following link:

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

Additional documentation can be found here:

 - National Weather Service: [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
 - National Climatic Data Center: [Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

#### Required packages for analysis

```{r Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr) # For manipulating data
library(ggplot2) # For generating visualisation
library(gridExtra) # For generating the bin width comparison plot
```

#### Reading and exploring the data frame

```{r reading data}
# Reading Storm dataset
if(!exists("dtStorm")) {
  stormDF <- read.csv("repdata_data_StormData.csv.bz2", header = TRUE, sep = ",", stringsAsFactors = FALSE)
}
```

```{r explore data}
str(stormDF) # Structure
```

#### Manipulating data for analysis 

Observations in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. Therefore, only the period from 1996 to 2011 has been considered for a more accurated analysis.*

```{r transforming and cleaning data}        
# Changing the class format
stormDF$BGN_DATE <- as.Date(stormDF$BGN_DATE, '%m/%d/%Y')
# Subsetting data as from 1996
new_stormDF <- stormDF[stormDF$BGN_DATE >= '1996-01-01',]
```

Storm database contains a total of 37 variables. To simplify the dataset further it has been subsetted to include only the variables interesting to our research questions. 
In our case, beside the *Event_type* and the *Date* columns, the variables *Fatalities* and *Injuries* has been used to analyse Health impacts while *Property damages* and *Crop damages* has been used to check the Economic effects of weather disasters. 

```{r Creating Health and Economic new_stormDF for analysis}
# Creating a table with Health data
stormHealthDF <- subset(new_stormDF, select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES))
names(stormHealthDF) <- c("Date", "Event_type", "Fatalities", "Injuries")
# Creating a table with Economic data 
stormEcoDF <- subset(new_stormDF, select=c(BGN_DATE, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
names(stormEcoDF) <- c("Date", "Event_type", "Property_Damages", "Property_Damages_Units", "Crop_Damages", "Crop_Damages_Units")
```

There are many observations with values equal to zero, therefore a filtering strategy has been made to avoid zero values in all rows. 

```{r filtering}
# Filtering data of both tables (Health and Economic) with values greater than 0 in all columns. 
stormHealthDF <- stormHealthDF[stormHealthDF$Injuries > 0 | stormHealthDF$Fatalities > 0  , c("Date", "Event_type", "Fatalities", "Injuries")]
stormEcoDF <- stormEcoDF[stormEcoDF$Property_Damages > 0 | stormEcoDF$Crop_Damages > 0  , c("Date", "Event_type", "Property_Damages", "Property_Damages_Units", "Crop_Damages", "Crop_Damages_Units")]
```

After subssetting the Storm database by a period of time, choosing only the columns we choose for analysis and filtering data greater than zero, we still see far more type of events (EVTYPE) recorded than the official 48. To achieve the 48 offical weather-related type of events we need to expend a lot of time cleaning the dataset. In the time limitation of this analysis, the "toupper" function has been used to reduce the "event types" in the dataset, which converts a lowercase string to a uppercase string. **Note**: other strategies and cleaning producedures should be done to perform a better analysis. Check further data analysis: unique(stormHealthDF$Event_type).

```{r toupper}
# Converting lower case to upper case
stormHealthDF$Event_type <- toupper(stormHealthDF$Event_type)
stormEcoDF$Event_type <- toupper(stormEcoDF$Event_type)
stormEcoDF$Property_Damages_Units <- toupper(stormEcoDF$Property_Damages_Units)
stormEcoDF$Crop_Damages_Units <- toupper(stormEcoDF$Crop_Damages_Units)
```

In order to analyse **which storm events are the most harmful to population health in the US**, it has been selected the most harmful weather events that caused the highest number of deaths and injuries among the US population. First by grouping data per type of the weather event and subsetting only the most harmful for the health society. 

```{r Summarising health variables by type of weather event}
# Create a table for Fatalities
agg_healthFatalities <- aggregate(Fatalities~Event_type, stormHealthDF, sum)
# Rename columns
names(agg_healthFatalities) <- c("Event_Type", "Total_Fatalities")
# Create a table for Injuries
agg_healthInjuries <- aggregate(Injuries~Event_type, stormHealthDF, sum)
# Rename columns
names(agg_healthInjuries) <- c("Event_Type", "Total_Injuries")
# Select the most severe Weather Events impacts in Health
top_healthFatalities <- agg_healthFatalities[order(agg_healthFatalities$Total_Fatalities, decreasing = TRUE), ][1:10, ]
top_healthInjuries <- agg_healthInjuries[order(agg_healthInjuries$Total_Injuries, decreasing = TRUE), ][1:10, ]
# Check the databases
top_healthFatalities
top_healthInjuries
```

## Results

Below, it can be visualised the highest effects on health and the economy caused by severe weather events in he United States between 1996 and 2011.

```{r Health impact plots}
# Plot Death plot and store in variable
ggplot(top_healthFatalities, aes(reorder(Event_Type, -Total_Fatalities), Total_Fatalities, fill=Event_Type)) + 
        geom_bar(stat= "identity", show.legend=F) +
        geom_text(aes(label=Total_Fatalities), vjust=1.6, color="black", position=position_dodge(0.9), size=2.5) +
        theme(axis.text.x=element_text(angle=45, hjust=1)) +
        xlab("Weather Events") + ylab("Number of Deaths") + 
        ggtitle("Highest fatalities caused by severe Weather Events in U.S. (1996 - 2011)")
       
# Plot Injuries plot and store in variable
ggplot(top_healthInjuries, aes(reorder(Event_Type, -Total_Injuries), Total_Injuries, fill=Event_Type)) + 
        geom_bar(stat= "identity", show.legend=F) +
        geom_text(aes(label=Total_Injuries), vjust=1.6, color="black", position=position_dodge(0.9), size=2.5) +
        theme(axis.text.x=element_text(angle=45, hjust=1)) +
        xlab("Weather Events") + ylab("Number of Injuries") + 
        ggtitle("Highest injuries caused by severe Weather Events in U.S. (1996 - 2011)")
      
```

Figure 1 and figure 2 highlights that excessive heat and tornadoes caused the highest number of fatalaties between 1996-2011, while tornadoes caused the greatest number of injuries.

And finally to asses **which types of extreme weather events have the greatest economic consequences in the U.S.** it has been selected the most harmful weather events that caused the highest number of damages in properties and agriculture in the US.  


```{r Summarising economic variables by type of weather event}
# Create two tables. One for Property and the other for Crop Damages
agg_eccProp <- aggregate(stormEcoDF$Property_Damages, by=list(stormEcoDF$Event_type), FUN="sum")
agg_eccCrop <- aggregate(stormEcoDF$Crop_Damages, by=list(stormEcoDF$Event_type), FUN="sum")
# Rename columns
names(agg_eccProp) <- c("Event_Type", "Total_Property_Damage")
names(agg_eccCrop) <- c("Event_Type", "Total_Crop_Damage")
# Select the most severe Weather Events impacts in the Economy
top_eccProp <- agg_eccProp[order(agg_eccProp$Total_Property_Damage, decreasing = TRUE), ][1:10, ]
top_eccCrop <- agg_eccCrop[order(agg_eccCrop$Total_Crop_Damage, decreasing = TRUE), ][1:10, ]
# Check the databases
top_eccProp
top_eccCrop
```
According to the National Weather Service instruction, estimates values has three significant digits followed by an alphabetical character signifying the magnitude of the number ("K" for thousands, "M" for millions, and "B" for billions). Therefore, Before ploting the histograms of economic impacts, values has been converted into the same numerical values. 

```{r magnitudes, message=FALSE, warning=FALSE}
# Converting all values into the same unit (large numbers)
top_eccProp$Total_Property_Damages[top_eccProp$Total_Property_Damages_Units == "K"] <- top_eccProp$Total_Property_Damages*1000 # Thousands
top_eccProp$Total_Property_Damages[top_eccProp$Total_Property_Damages_Units == "M"] <- top_eccProp$Total_Property_Damages*1000000 # Million
top_eccProp$Total_Property_Damages[top_eccProp$Total_Property_Damages_Units == "B"] <- top_eccProp$Total_Property_Damages*1000000000 # Billion

top_eccCrop$Total_Crop_Damages[top_eccCrop$Total_Crop_Damages_Units == "K"] <- top_eccCrop$Total_Crop_Damages*1000 # Thousands
top_eccCrop$Total_Crop_Damages[top_eccCrop$Total_Crop_Damages_Units == "M"] <- top_eccCrop$Total_Crop_Damages*1000000 # Million
top_eccCrop$Total_Crop_Damages[top_eccCrop$Total_Crop_Damages_Units == "B"] <- top_eccCrop$Total_Crop_Damages*1000000000 # Billion

head(top_eccProp)
```

```{r Economic impact plots}
# Plotting Property Damages 
p1 <- ggplot(top_eccProp, aes(reorder(Event_Type, -Total_Property_Damage), round(Total_Property_Damage/1000000,2), fill=Event_Type)) + 
        geom_bar(stat= "identity", show.legend=F) +
        geom_text(aes(label=round(Total_Property_Damage/1000000,2)), vjust=1.6, color="black", position=position_dodge(0.9), size=2.5) +
        theme(axis.text.x=element_text(angle=45, hjust=1)) +
        xlab("Weather Events") + ylab("Total Damage in Million Dolars") + 
        ggtitle("Impacts in properties caused by Weather Events (US 1996 - 2011)") + theme(plot.title = element_text(size = 10)) +
        ylim(0.0, 1.5)
       
# Plotting Crop Damages 
p2 <- ggplot(top_eccCrop, aes(reorder(Event_Type, -Total_Crop_Damage), round(Total_Crop_Damage/1000000,2), fill=Event_Type)) + 
        geom_bar(stat= "identity", show.legend=F) +
        geom_text(aes(label=round(Total_Crop_Damage/1000000,2)), vjust=1.6, color="black", position=position_dodge(0.9), size=2.5) +
        theme(axis.text.x=element_text(angle=45, hjust=1)) +
        xlab("Weather Events") + ylab("Total Damage in Million Dolars") + 
        ggtitle("Impacts in agriculture caused by Weather Events (US 1996 - 2011)") + theme(plot.title = element_text(size = 10)) +
        ylim(0.0, 1.5)

grid.arrange(p1,p2, layout_matrix = rbind(c(6, 1.2),c(6, 1.2)))

```

Figure 3 shows that Thunderstorm Wind, Flash Flood and Tornadoes have been devastating and have caused majority of the property damage amounting to approximatelly 1.3 million dollars of public property. On the other hand, Hail account for majority crop damages across the US amounting to 0.5 Million Dollars.

## Conclusion:

Based on the plots we can say that Tornadoes, and other events such as Floods have been the major event to cause maximum fatalities and Injuries, while Thunderstorm Winds, Flash Flood and Tornadoes have been economically devastating throught the period from 1996 to 2011.
